---
title: "R Notebook"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
---


<h1><center>Crimes in Chicago - Differences across years 2006 and 2016</center></h1>

<br>

<h4>Aims</h4>
<p>Put objective here!!! </p> 


```{r include=FALSE}

#### Libraries

## Load libraries
library(readr)
suppressPackageStartupMessages(library(dplyr))
library(tidyr)
suppressPackageStartupMessages(library(lubridate))
library(ggplot2)
suppressPackageStartupMessages(library(highcharter))
suppressPackageStartupMessages(library(viridis))
suppressPackageStartupMessages(library(gridExtra))


#### Get raw data

## Set working directory
setwd("C:/Users/Muhsin Karim/Documents/Data/raw_data/kaggle/crimes_in_chicago")

## Load in each file
df1 <- read_csv("Chicago_Crimes_2001_to_2004.csv")
df2 <- read_csv("Chicago_Crimes_2005_to_2007.csv")
df3 <- read_csv("Chicago_Crimes_2008_to_2011.csv")
df4 <- read_csv("Chicago_Crimes_2012_to_2017.csv")

## Bind rows
df <- rbind(df1, df2, df3, df4)
rm(df1, df2, df3, df4)

## Remove incomplete years
df <- df[!(df$Year %in% c("2001", "2017")), ]


### Get Community Area name
# Data from: https://data.cityofchicago.org/Facilities-Geographic-Boundaries/Boundaries-Community-Areas-current-/cauq-8yn6

dfCommunity <- read_csv("CommAreas.csv")
dfCommunity <- dfCommunity %>% 
  select(`Community Area Code` = AREA_NUMBE, 
         `Community Area` = COMMUNITY)
colnames(df)[colnames(df) == "Community Area"] <- "Community Area Code"
df <- left_join(df, dfCommunity, by = "Community Area Code")

## Create a table mapping District to Community Area
district_community <- na.omit(df) %>% select(District, `Community Area`) %>% distinct(District, `Community Area`) %>% arrange(District, `Community Area`)


#### Dates and time

## Create datetimes
df$Datetime <- as.POSIXct(df$Date, format="%m/%d/%Y %I:%M:%S %p")
df$`Updated On` <- as.POSIXct(df$`Updated On`, format="%m/%d/%Y %I:%M:%S %p")


## Create datetime parts
df$Day <- factor(day(df$Datetime))
df$Month <- factor(month(df$Datetime, label = TRUE))
df$Year <- factor(year(df$Datetime))
df$Weekday <- factor(wday(df$Datetime, label = TRUE))
df$Hour <- factor(hour(df$Datetime))

## Create date
df$Date <- as.Date(df$Datetime)


#### Remove duplicates

## Create unique identifier
df$Identifier <- paste(df$ID, df$`Case Number`, sep = "-")

## Remove duplicates based on ID and Case Number
df <- df[order(df$`Updated On`, decreasing = T), ]
if (sum(duplicated(df$Identifier)) > 0) {df <- df[-which(duplicated(df$Identifier)), ]}


#### Create timeseries

# ## By Date
# by_date <- na.omit(df) %>% group_by(Date) %>% summarise(Total = n())
#tseries_date <- xts(by_date$Total, order.by = as.POSIXct(by_date$Date))

## By Year
by_year <- na.omit(df) %>% group_by(Year) %>% summarise(Total = n()) %>% arrange(Year)


#### All crimes

### Percent change across years

## Group and get percent change
by_year$Previous <- lag(by_year$Total, 1, na.pad = TRUE)
by_year$`Percent Change` <- round((by_year$Total - by_year$Previous) / by_year$Total * 100)

```

```{r echo=FALSE}

## Plot number of crimes per year
hchart(by_year, "column", hcaes(x = Year, y = Total, color = Year)) %>%
  hc_add_theme(hc_theme_google()) %>%
  hc_title(text = "Number of Crimes by Year") %>%
  hc_credits(enabled = TRUE, text = "Sources: Chicago Police Department", style = list(fontSize = "12px")) %>%
  hc_legend(enabled = FALSE)

```

<br>

The number of crimes peaked in 2003 (470,958 crimes) then decreased since 2005. I would like to investigate the differences in crime across years, namely the percent increase or decrease in crime and the possible reasons that account for changes given the data. Here is the percent change between each year.

<br>
<br>


```{r echo=FALSE}

## Plot percent change
hchart(by_year, "column", hcaes(x = Year, y = `Percent Change`, color = Year)) %>%
  hc_add_theme(hc_theme_google()) %>%
  hc_title(text = "Crime Percent Change per Year") %>%
  hc_credits(enabled = TRUE, text = "Sources: Chicago Police Department", style = list(fontSize = "12px")) %>%
  hc_legend(enabled = FALSE)

```

<br>

Since 2006, the crime percent change from previous years has decreased. There are large swings in early years. The number of crimes increased from 2002 to 2003 by 27% (bar at the '2003' x-axis label). It decreased by 22% from 2003 and 2004, then increased by 14% from 2004 to 2005. What accounts for these swings? I suspect it is due to data entry/record keeping as opposed to a dramatic change in crime in early 2000s. I discuss this at the end - see "Crime from 2002 to 2004".


Let's look at specific crime types starting with theft.

<br>
<br>


```{r include=FALSE}

##### Crime between 2006 and 2016

### Create data frame with select years

dfTen <- df[df$Year %in% c("2006", "2016"), ] 


#### Functions----

### District line plot

plot_district <- function(title) {
  ggplot(data = by_district, aes(x = Year, y = Total, group = District, colour = as.factor(District))) +
    geom_line(size = 1.5) +
    guides(colour = guide_legend(title = "District")) +
    scale_x_discrete(expand = c(-0.25, 0.25)) +
    ggtitle(title) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
}


### Communtity line plot

plot_community <- function(title) {
  ggplot(data = by_community, aes(x = Year, y = Total, group = `Community Area`, colour = as.factor(`Community Area`))) +
    geom_line(size = 1) +
    scale_x_discrete(expand = c(-0.25, 0.25)) +
    ggtitle(title) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5)) +
    theme(legend.position = "none")
}


### Year and Month heatmap

plot_year_month <- function(title_2006, title_2016) {
  
  ## 2006 
  period_counts1 <- na.omit(dfCrime) %>% filter(Year == "2006") %>% group_by(Year, Month) %>% summarise(Total = n())
  gg_period1 <- ggplot(period_counts1, aes(Year, Month, fill = Total)) +
    geom_tile(size = 1, color = "white") +
    scale_fill_viridis()  +
    geom_text(aes(label=Total), color='white') +
    ggtitle(title_2006)

  ## 2016 
  period_counts2 <- na.omit(dfCrime) %>% filter(Year == "2016") %>% group_by(Year, Month) %>% summarise(Total = n())
  gg_period2 <- ggplot(period_counts2, aes(Year, Month, fill = Total)) +
    geom_tile(size = 1, color = "white") +
    scale_fill_viridis()  +
    geom_text(aes(label=Total), color='white') +
    ggtitle(title_2016)
  
  ## Plot both
  grid.arrange(gg_period1, gg_period2, ncol = 2)
}


### Weekday and Hour heatmap

plot_weekday_hour <- function(title_2006, title_2016) {
  
  ## 2006
  period_counts1 <- na.omit(dfCrime) %>% filter(Year == "2006") %>% group_by(Weekday, Hour) %>% summarise(Total = n())
  gg_period1 <- ggplot(period_counts1, aes(Weekday, Hour, fill = Total)) +
    geom_tile(size = 1, color = "white") +
    scale_fill_viridis()  +
    geom_text(aes(label=Total), color='white') +
    ggtitle(title_2006)
  
  ## 2016
  period_counts2 <- na.omit(dfCrime) %>% filter(Year == "2016") %>% group_by(Weekday, Hour) %>% summarise(Total = n())
  gg_period2 <- ggplot(period_counts2, aes(Weekday, Hour, fill = Total)) +
    geom_tile(size = 1, color = "white") +
    scale_fill_viridis()  +
    geom_text(aes(label=Total), color='white') +
    ggtitle(title_2016)
  
  ## Plot both
  grid.arrange(gg_period1, gg_period2, ncol = 2)
}


### District table

table_district <- function() {
  by_district_spread <- by_district %>% spread(Year, Total)
  by_district_spread$`Percent Change` <- round(as.numeric(unlist((by_district_spread[ , 3] - by_district_spread[ , 2]) / by_district_spread[ , 3] * 100)))
  by_district_spread <- by_district_spread %>% arrange(`Percent Change`)
  by_district_spread 
}


### Community table

table_community <- function() {
  by_community_spread <- by_community %>% spread(Year, Total)
  by_community_spread$`Percent Change` <- round(as.numeric(unlist((by_community_spread[ , 3] - by_community_spread[ , 2]) / by_community_spread[ , 3] * 100)))
  by_community_spread <- by_community_spread %>% arrange(`Percent Change`)
  by_community_spread
}

```


```{r include=FALSE}

#### Theft 2006 and 2016----

### Subset by crime type

dfCrime <- dfTen[(dfTen$`Primary Type` == "THEFT"), ]


### Districts

## Group by District
by_district <- na.omit(dfCrime) %>% group_by(Year, District) %>% summarise(Total = n()) %>% arrange(Year)### Year by month

```

```{r echo=FALSE}

## Plot line graph
plot_district(title="Numbers of Theft per District")

```

<br>

Essentially all Districts experienced a decrease in theft. The table below shows the percent decreased in crime per District.

<br>
<br>

```{r echo=FALSE}

## Display table
table_district()

```

```{r echo=FALSE}

### Community Areas

## Group by Community Area
by_community <- na.omit(dfCrime) %>% group_by(Year, `Community Area`) %>% summarise(Total = n()) %>% arrange(Year)### Year by month

## Plot line graph
plot_community(title="Numbers of Theft per Community")

```

<br>

Not all Community Areas experienced a drop in theft. The table below shows that Avalon Park had the greatest decrease in theft (154%). Oakland experiences a 41 % increased in theft. In Oakland's defence, I do not have the population numbers in 2006 and 2016. Perhaps there was a significant increase in the residence of Oaklands that accounts for part of the increase in theft. 

```{r echo=FALSE}

## Display table
table_community()

```

```{r echo=FALSE}

### Year by Month

## Plot heatmap
plot_year_month(title_2006="Numbers of Theft in 2006", title_2016="Numbers of Theft in 2016")

```

<br>

The numbers of theft are proportional across the months over both years. The summer months are a time for theft in Chicago.

<br>
<br>


```{r echo=FALSE}

### Weekday by Hour

## Plot heatmap
plot_weekday_hour(title_2006="Numbers of Theft in 2006", title_2016="Numbers of Theft in 2016")

```

<br>
<br>

Again, theft is proportional across both years per weekday and hour. There does appear to be a theft hotspot around weekdays at noon during 2006 (possible because residents are at work, leaving homes vacant). In 2016, the hotspots are on Thursday and Fridays at 6 pm.   

