---
output: html_document
---

<h1><center> Chicago Crime Report (2012-2016)</center></h1>

<br>


<h4> Crime in Chicago</h4>
<p> Crime in Chicago has been tracked by the Chicago Police Department's Bureau of Records since the beginning of the 20th century. The city's overall crime rate, especially the violent crime rate, is substantially higher than the US average. Chicago was responsible for nearly half of 2016's increase in homicides in the US, though national crime rates stayed near historic lows.As of 2017, Chicago's homicide rate is significantly higher when compared to the larger American cities of New York and Los Angeles, but lower when compared to smaller American cities. The reason for the violence which is localized to some areas of the city, including change in police tactics or increase in gang rivalry, remain unclear. </p> 



<br>
<h4>Why is this dataset more important now?</h4>

* Chicago Crime Numbers have been in the media for all the wrong reasons. 
* President widely used these numbers during his presidential campaign
* He still talks about it.
* These numbers are impacting the overall numbers of the country.
* Did homicides rates really getting worse in Chicago in 2016?



```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, fig.width = 6, fig.height = 5, fig.align = "center")

library(highcharter)
library(ggplot2)
library(dplyr)
library(tidyr)
library(viridis)
library(plotly)
library(lubridate)
library(xts)
library(maps)
library(ggmap)
library(gridExtra)

chicagocrimes20122016 <- read.csv("../input/Chicago_Crimes_2012_to_2017.csv")

## Reading the data
chicagocrimes20122016 <- chicagocrimes20122016[chicagocrimes20122016$Year!='2017',]


### Working with Date and Time
chicagocrimes20122016$Day <- factor(day(as.POSIXlt(chicagocrimes20122016$Date, format="%m/%d/%Y %I:%M:%S %p")))
chicagocrimes20122016$Month <- factor(month(as.POSIXlt(chicagocrimes20122016$Date, format="%m/%d/%Y %I:%M:%S %p"), label = TRUE))
chicagocrimes20122016$Year <- factor(year(as.POSIXlt(chicagocrimes20122016$Date, format="%m/%d/%Y %I:%M:%S %p")))
chicagocrimes20122016$Weekday <- factor(wday(as.POSIXlt(chicagocrimes20122016$Date, format="%m/%d/%Y %I:%M:%S %p"), label = TRUE))

chicagocrimes20122016$Date <- as.Date(chicagocrimes20122016$Date, "%m/%d/%Y %I:%M:%S %p")

## Creating timeseries
by_Date <- na.omit(chicagocrimes20122016) %>% group_by(Date) %>% summarise(Total = n())
tseries <- xts(by_Date$Total, order.by=as.POSIXct(by_Date$Date))


## Creating timeseries of arrests made
Arrests_by_Date <- na.omit(chicagocrimes20122016[chicagocrimes20122016$Arrest == 'True',]) %>% group_by(Date) %>% summarise(Total = n())
arrests_tseries <- xts(Arrests_by_Date$Total, order.by=as.POSIXct(by_Date$Date))

## BY Location
by_location <- chicagocrimes20122016 %>% group_by(Location.Description) %>% summarise(Total = n()) %>% arrange(desc(Total))

## By Primary Type
by_type <- chicagocrimes20122016 %>% group_by(Primary.Type) %>% summarise(Total = n()) %>% arrange(desc(Total))
  

## By District
by_district <- chicagocrimes20122016 %>% group_by(District) %>% summarise(Total = n()) %>% arrange(desc(Total))


## By Primary Type
by_ward <- chicagocrimes20122016 %>% group_by(Ward) %>% summarise(Total = n()) %>% arrange(desc(Total))


## By FBI Code
by_fbi <- chicagocrimes20122016 %>% group_by(FBI.Code) %>% summarise(Total = n()) %>% arrange(desc(Total))

## By Arrest
by_arrest <- chicagocrimes20122016 %>% group_by(Arrest) %>% summarise(Total = n()) %>% arrange(desc(Total))

## By Domestic
by_domestic <- chicagocrimes20122016 %>% group_by(Domestic) %>% summarise(Total = n()) %>% arrange(desc(Total))

## By Year
by_year <- chicagocrimes20122016 %>% group_by(Year) %>% summarise(Total = n()) %>% arrange(desc(Total))

## Lat and Long of Crimes
LatLonCounts <- as.data.frame(table(round(chicagocrimes20122016$Longitude,2), round(chicagocrimes20122016$Latitude,2)))
LatLonCounts$Long <- as.numeric(as.character(LatLonCounts$Var1))
LatLonCounts$Lat <- as.numeric(as.character(LatLonCounts$Var2))
LatLonCounts2 <- subset(LatLonCounts, Freq > 0)

## Lat and Long of Arrests
arrests_data <- na.omit(chicagocrimes20122016[chicagocrimes20122016$Arrest == 'True',])
LatLonArrestsCounts <- as.data.frame(table(round(arrests_data$Longitude,2), round(arrests_data$Latitude,2)))
LatLonArrestsCounts$Long <- as.numeric(as.character(LatLonArrestsCounts$Var1))
LatLonArrestsCounts$Lat <- as.numeric(as.character(LatLonArrestsCounts$Var2))
LatLonArrestsCounts2 <- subset(LatLonArrestsCounts, Freq > 0)

```


```{r}

hchart(tseries, name = "Crimes") %>% 
  hc_add_series(arrests_tseries, name = "Arrests") %>%
  hc_add_theme(hc_theme_darkunica()) %>%
  hc_credits(enabled = TRUE, text = "Sources: City of Chicago Administration and the Chicago Police Department", style = list(fontSize = "12px")) %>%
  hc_title(text = "Times Series plot of Chicago Crimes and Arrests") %>%
  hc_legend(enabled = TRUE)

```

<br>
<br>

* Crimes have decreased in Chicago in 2016 compared to 2012. 
* There is clear indication in the timeseries plot that crime numbers increase somewhere during middle of the year mostly during summer months and drops during end/starting of the year mostly during winter months.
* Why are the number of arrests so less compared to number of crimes?
* It would be interesting to look at how the crimes are distributed by months.

<br>
<br>


```{r}

hchart(arrests_tseries) %>%
  hc_add_theme(hc_theme_google()) %>%
  hc_title(text = "Times Series plot of Arrests made in Chicago (2012-2016)") %>%
  hc_credits(enabled = TRUE, text = "Sources: City of Chicago Administration and the Chicago Police Department", style = list(fontSize = "12px"))

```

<br>
<br>
 
* This plot clearly shows how much have the arrests decreased from 2012-2016
* It is surprising given the fact that there is lot of critisism about Chicago Crimes 

<br>
<br>


```{r fig.height=7, fig.width=9}


arrests_count <- arrests_data %>% group_by(Year, Month) %>% summarise(Total = n())

arrests <- ggplot(arrests_count, aes(Year, Month, fill = Total)) +
  geom_tile(size = 1, color = "white") +
  scale_fill_viridis()  +
  geom_text(aes(label=Total), color='white') +
  ggtitle("Arrests by Year and Month(2012-2016)")


crime_count <- chicagocrimes20122016 %>% group_by(Year, Month) %>% summarise(Total = n())

crimes <- ggplot(crime_count, aes(Year, Month, fill = Total)) +
  geom_tile(size = 1, color = "white") +
  scale_fill_viridis()  +
  geom_text(aes(label=Total), color='white') +
  ggtitle("Crimes by Year and Month(2012-2016)")


grid.arrange(crimes, arrests, ncol = 2)

```


<br>

* Heatmap clearly shows how the number of arrests have decreased by more than a half between 2012 and 2016 but the crimes have not reduced at the same rate. Still the arrests have gone done drastically.

<br>

<!-- ```{r} -->

<!-- hchart(by_year, "column", hcaes(x = Year, y = Total, color = Year)) %>% -->
<!--   hc_colorAxis(stops = color_stops(n = 10, colors = c("#440154", "#21908C", "#FDE725"))) %>% -->
<!--   hc_add_theme(hc_theme_darkunica()) %>% -->
<!--   hc_title(text = "Number of Crimes by Year") %>% -->
<!--   hc_credits(enabled = TRUE, text = "Sources: City of Chicago Administration and the Chicago Police Department", style = list(fontSize = "12px")) %>% -->
<!--   hc_legend(enabled = FALSE) -->

<!-- ``` -->

<br>
<br>


```{r}

hchart(by_location[1:20,], "column", hcaes(x = Location.Description, y = Total, color = Total)) %>%
  hc_colorAxis(stops = color_stops(n = 10, colors = c("#440154", "#21908C", "#FDE725"))) %>%
  hc_add_theme(hc_theme_darkunica()) %>%
  hc_title(text = "Locations with most Crimes - Top 20") %>%
  hc_credits(enabled = TRUE, text = "Sources: City of Chicago Administration and the Chicago Police Department", style = list(fontSize = "12px")) %>%
  hc_legend(enabled = FALSE)

```

<br>
<br>

* Streets are the most common location where crimes happen.
* With Apartments and Residence being other top common locations 

<br>
<br>

#### Let us understand how crime numbers vary in some of these top crime locations 

<br>
<br>

```{r}

streets <- chicagocrimes20122016[chicagocrimes20122016$Location.Description=="STREET",]
## Creating timeseries
streets_by_Date <- na.omit(streets) %>% group_by(Date) %>% summarise(Total = n())
streets_tseries <- xts(streets_by_Date$Total, order.by=as.POSIXct(by_Date$Date))

residence <- chicagocrimes20122016[chicagocrimes20122016$Location.Description=="RESIDENCE",]
## Creating timeseries
residence_by_Date <- na.omit(residence) %>% group_by(Date) %>% summarise(Total = n())
residence_tseries <- xts(residence_by_Date$Total, order.by=as.POSIXct(by_Date$Date))

apartment <- chicagocrimes20122016[chicagocrimes20122016$Location.Description=="APARTMENT",]
## Creating timeseries
apartment_by_Date <- na.omit(apartment) %>% group_by(Date) %>% summarise(Total = n())
apartment_tseries <- xts(apartment_by_Date$Total, order.by=as.POSIXct(by_Date$Date))

sidewalk <- chicagocrimes20122016[chicagocrimes20122016$Location.Description=="SIDEWALK",] 
## Creating timeseries
sidewalk_by_Date <- na.omit(sidewalk) %>% group_by(Date) %>% summarise(Total = n())
sidewalk_tseries <- xts(sidewalk_by_Date$Total, order.by=as.POSIXct(by_Date$Date))

hchart(streets_tseries, name = "Streets") %>% 
  hc_add_series(residence_tseries, name = "Residence") %>% 
  hc_add_series(apartment_tseries, name = "Apartment") %>%
  hc_add_series(sidewalk_tseries, name = "Sidewalk") %>%
  hc_add_theme(hc_theme_darkunica()) %>%
  hc_credits(enabled = TRUE, text = "Sources: City of Chicago Administration and the Chicago Police Department", style = list(fontSize = "12px")) %>%
  hc_title(text = "Crimes in Streets/Residence/Apartment/Sidewalk") %>%
  hc_legend(enabled = TRUE)

```


<br>
<br>

* There is definite reduction in crimes in top crime locations
* Particularly in sidewalks there is drastic reduction.

<br>
<br>


```{r}

hchart(by_type, "column", hcaes(Primary.Type, y = Total, color = Total)) %>%
  hc_colorAxis(stops = color_stops(n = 10, colors = c("#440154", "#21908C", "#FDE725"))) %>%
  hc_add_theme(hc_theme_darkunica()) %>%
  hc_title(text = "Crime Types") %>%
  hc_credits(enabled = TRUE, text = "Sources: City of Chicago Administration and the Chicago Police Department", style = list(fontSize = "12px")) %>%
  hc_legend(enabled = FALSE)

```


<br>
<br>

* Thefts being the most common crimes followed by Battery.
* What is Battery? 
<p>A person may be charged with the crime of battery under Illinois law if he or she makes actual physical contact with another individual with the intent to injure, provoke, or insult that person. A simple battery charge may result in a fine of up to $2,500.00 and a jail sentence of up to a year. </p>

<br>
<br>

#### Let us see how crime numbers vary in top crimes

<br>
<br>

```{r}

thefts <- chicagocrimes20122016[chicagocrimes20122016$Primary.Type=="THEFT",] 
## Creating timeseries
thefts_by_Date <- na.omit(thefts) %>% group_by(Date) %>% summarise(Total = n())
thefts_tseries <- xts(thefts_by_Date$Total, order.by=as.POSIXct(by_Date$Date))

battery <- chicagocrimes20122016[chicagocrimes20122016$Primary.Type=="BATTERY",] 
## Creating timeseries
battery_by_Date <- na.omit(battery) %>% group_by(Date) %>% summarise(Total = n())
battery_tseries <- xts(battery_by_Date$Total, order.by=as.POSIXct(by_Date$Date))

criminals <- chicagocrimes20122016[chicagocrimes20122016$Primary.Type=="CRIMINAL DAMAGE",]
## Creating timeseries
criminals_by_Date <- na.omit(criminals) %>% group_by(Date) %>% summarise(Total = n())
criminals_tseries <- xts(criminals_by_Date$Total, order.by=as.POSIXct(by_Date$Date))

narcotics <- chicagocrimes20122016[chicagocrimes20122016$Primary.Type=="NARCOTICS",] 
## Creating timeseries
narcotics_by_Date <- na.omit(narcotics) %>% group_by(Date) %>% summarise(Total = n())
narcotics_tseries <- xts(narcotics_by_Date$Total, order.by=as.POSIXct(by_Date$Date))


hchart(thefts_tseries, name = "Thefts") %>% 
  hc_add_series(battery_tseries, name = "Battery") %>% 
  hc_add_series(criminals_tseries, name = "Criminal Damage") %>%
  hc_add_series(narcotics_tseries, name = "Narcotics") %>%
  hc_add_theme(hc_theme_darkunica()) %>%
  hc_credits(enabled = TRUE, text = "Sources: City of Chicago Administration and the Chicago Police Department", style = list(fontSize = "12px")) %>%
  hc_title(text = "Crimes in Thefts/Battery/Criminal Damage/Narcotics") %>%
  hc_legend(enabled = TRUE)


```


<br>
<br>

* Number of Narcotics crimes have reduced.
* Number of Thefts and Battery crimes have remained the same.

<br>
<br>

### How big is the increase in homicides in Chicago?


```{r}

homicide <- chicagocrimes20122016[chicagocrimes20122016$Primary.Type=="HOMICIDE",] 

homicide_year <-  homicide %>% group_by(Year) %>% summarise(Total = n())

hchart(homicide_year, "column", hcaes(Year, Total, color = Year)) %>%
  hc_add_theme(hc_theme_darkunica()) %>%
  hc_title(text = "Homicide 2012-2016")  %>%
  hc_credits(enabled = TRUE, text = "Sources: City of Chicago Administration and the Chicago Police Department", style = list(fontSize = "12px"))

```

<br>
<br>

```{r}


homicide_count <- homicide %>% group_by(Year, Month) %>% summarise(Total = n())

ggplot(homicide_count, aes(Year, Month, fill = Total)) +
  geom_tile(size = 1, color = "white") +
  scale_fill_viridis()  +
  geom_text(aes(label=Total), color='white') +
  ggtitle("Homicides in Chicago (2012-2016)")

```


<br>
<br>


* There is huge increase in the number of homicides in Chicago in 2016 compared to previous years.
* Shocking that authorities are clueless about the reasons 
* It is worrying how the people in power are using the same for publicity when they should be sitting and talking on how to solve the problem at hand.
* Being one of the biggest cities it does have an impact on the overall crime numbers of the country. Hence the talk about Chicago crime when lot of other smaller cities have bigger crime rates.


#### References
* https://en.wikipedia.org/wiki/Battery_(crime)
* https://en.wikipedia.org/wiki/Crime_in_Chicago
* https://www.theatlantic.com/politics/archive/2017/01/chicago-homicide-spike-2016/514331/




